Free surface computations in mantle convection models

### 2016-04-04 free-surface-paper.tex
Add Cedric to acknowledgements
Address review one comments.

\begin{abstract}
Geodynamic simulations increasingly rely on simulations with a true free surface to 
investigate questions of dynamic topography, tectonic deformation, gravity perturbations, and
global mantle convection. However, implementations of free surface boundary conditions 
have proven challenging from a standpoint of accuracy, robustness, and stability.
In particular, free surfaces tend to suffer from sloshing instabilities, also known as 
the ``drunken sailor'' instability, which severely limit time step sizes. Several 
schemes have been proposed in the literature to deal with these instabilities.

@@

Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit time stepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{kramer2012implicit} and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.

@@

We begin with the incompressible momentum conservation equations for creeping incompressible flow:
\begin{equation}
\begin{aligned}
\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
\nabla \cdot \mathbf{u} &= 0
\end{aligned}
\label{eq:momentum_mass_conservation}
\end{equation}
where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
\label{eq:stress_tensor}
\end{equation}
where $\eta$ is the viscosity and $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$ is the strain-rate tensor.
Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}
\label{eq:stokes}
\end{equation}

@@

Formally, this gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad &\text{in} \qquad &\Omega(t)\\
\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in} \qquad &\Omega(t)
\end{aligned}
\label{eq:final_system}
\end{equation}
defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
For the sake of economy we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{u}_{\mathrm{mesh}}(t)$
\begin{equation}
 \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}_{\mathrm{mesh}}(t) \text{ d}t.
\end{equation}
The mesh velocity in normal direction is given by the velocity $\mathbf{u}(t)$ on the free surface
\begin{equation}
 \mathbf{u}_\mathrm{mesh}(t) \cdot \mathbf{n} = \mathbf{u}(t) \cdot \mathbf{n} \quad \text{on} \quad \Gamma_F
\end{equation}
and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{u}_\mathrm{mesh}$ in the interior of the domain
to Section~\ref{sec:remeshing}.
Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{mesh}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
equation
\begin{equation}
\frac{\text{d} \zeta}{\text{d}t} = \mathbf{u \cdot \mathbf{n}} \quad \textrm{on  }  \Gamma_F.
\label{eq:surface_evolution}
\end{equation}

@@

Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
is not invertible. It may, however, be transformed into a standard eigenvalue problem.
Define
\begin{equation}
\begin{aligned}
C &= 
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix} \quad
D &= 
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix} \quad
\mathbf{y}_i &= 
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix} 
\end{aligned}
\end{equation}
Then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
\begin{equation}
(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i
\label{eq:standard_eigenvalue}
\end{equation}

@@

Armed with the normal modes and relaxation times of the Stokes system, we can write down the
formal solution to Equation~\ref{eq:surface_evolution}. Let the initial surface topography be 
represented by a linear combination of its normal modes
\begin{equation}
\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x})
\end{equation}
then the time evolution of the free surface is given by
\begin{equation}
\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
\end{equation}
Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
the solution and surface topography into its normal modes and integrate those separately. 
Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
The normal mode solution is instructive, however. 
Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
The decay equation is the archetypical example of a stiff ordinary differential equation.
If we were to numerically integrate this in time with a forward Euler method, we would find the 
time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
\begin{equation}
\Delta t  \le 2 \tau_{\mathrm{min}}.
\label{eq:cfl_euler}
\end{equation}
The maximum allowable timestep is limited by the minimum relaxation timescale.
If a larger timestep than this is taken then those modes will go unstable.
The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
and so it will be those which go unstable first, a phenomenon which has been called 
a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.

@@

Again we may take one of the normal modes as an example and inspect the difference between the nonstandard finite-difference
scheme and the analytical solution after one time step:

@@

\begin{itemize}
\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
only the minimun relaxation time is required. For certain geometries a simple power iteration on 
the standard eigenvalue problem~\eqref{eq:standard_eigenvalue} could be enough.
\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosty
variations), an analytical approximation may be sufficient.
\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
where the density, gravity, viscosity, and lengthscale are all representative values.  A small amount
of experimentation near this value of $\tau$ can find an appropriate relaxation time.
\item Observation of instabilities.  The mere act of observing a sloshing instability in an unstabilized
problem can furnish an estimate of its relaxation time.
\end{itemize}

@@

\subsection{Surface advection}
With a deformable domain there is the danger that small errors in free surface motion can
result in poor overall mass conservation in time. In some scenarios, the total volume of the mesh can 
fluctuate significantly over hundreds or thousands of time steps.
Consistency with the Stokes solution requires 
\begin{equation}
\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}.
\end{equation}

@@

This solution is only valid for infinitesimal topography. However, for small
initial topography $\zeta_0$ it seems to be sufficiently accurate to test convergence orders up 
to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
We evaluate error by time-integrating the L2 difference between the numerical and analytical
solutions at the center of the domain:
\begin{equation}
\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
\end{equation}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/amr.pdf}
\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement. The timestep $\Delta t$ is 500 yr. We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m). For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients. Both cases converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
\label{fig:amr}
\end{figure*}

@@

Finally, we have described the implementation of free surface boundary conditions in 
the open source mantle convection software \texttt{ASPECT}. Both the quasi-implicit 
scheme and the nonstandard finite difference scheme are available. The implementation is 
sufficiently general to accommodate many different geometies, rheologies, and 
gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.

\section*{Acknowledgements}
This work was supported through National Science Foundation grant EAR-1246670
and by the Computational Infrastructure in Geodynamics (CIG), through National Science Foundation grant EAR-0949446. 
TH is partially supported by NSF grant DMS-1522191.
Thanks to Wolfgang Bangerth and Brent Delbridge for useful discussions.
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

>>>>

\begin{abstract}
Geodynamic simulations increasingly rely on simulations with a true free surface to 
investigate questions of dynamic topography, tectonic deformation, gravity perturbations, and
global mantle convection. However, implementations of free surface boundary conditions 
have proven challenging from a standpoint of accuracy, robustness, and stability.
In particular, free surfaces tend to suffer from sloshing instabilities, also known as 
the ``drunken sailor'' instability, which severely limit timestep sizes. Several 
schemes have been proposed in the literature to deal with these instabilities.

@@

Several studies have attempted to alleviate the timestepping requirements imposed by the sloshing 
instability. Since the most expensive part of geodynamic simulations is typically the Stokes solve,
most free surface calculations have preferred to use explicit timestepping methods for the free 
surface. \citet{kaus2010stabilization} proposed a quasi-implicit scheme which modifies the discretized 
Stokes matrix, giving it better stability properties.  \citet{kramer2012implicit} and \citet{furuichi2015implicit}
explored methods for solving the transport of the free surface implicitly.

@@

We begin with the incompressible momentum conservation equations for creeping incompressible flow:
\begin{equation}
\begin{aligned}
\nabla \cdot \mathbf{T} &= \rho \mathbf{g} \\
\nabla \cdot \mathbf{u} &= 0
\end{aligned}
\label{eq:momentum_mass_conservation}
\end{equation}
where $\mathbf{u}$ is the fluid velocity, $\rho$ is the fluid density, and $\mathbf{g}$ is the force due to gravity.
$\mathbf{T}$ is the stress tensor for a Newtonian fluid, defined by
\begin{equation}
\mathbf{T} = 2 \eta \varepsilon(\mathbf{u}) - p \mathbf{I}
\label{eq:stress_tensor}
\end{equation}
where $\eta$ is the viscosity, $\varepsilon(\mathbf{u}) = \frac{1}{2}(\nabla \mathbf{u} + (\nabla \mathbf{u} )^T )$
is the strain-rate tensor, and $\mathbf{I}$ is the identity tensor.
Substituting the stress tensor into Equation~\eqref{eq:momentum_mass_conservation} gets the familiar form of the Stokes equation:
\begin{equation}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p = \rho \mathbf{g}
\label{eq:stokes}
\end{equation}

@@

Formally, this gives rise to the following time dependent, coupled system with unknowns $\mathbf{u}(t)$ and $\mathbf{p}(t)$:
\begin{equation}
\begin{aligned}
\nabla \cdot \left( 2 \eta \varepsilon( \mathbf{u} ) \right) - \nabla p &= \rho \mathbf{g} \qquad &\text{in} \qquad &\Omega(t)\\
\nabla \cdot \mathbf{u} &= 0  \qquad &\text{in} \qquad &\Omega(t)
\end{aligned}
\label{eq:final_system}
\end{equation}
defined on the bounded, moving domain $\Omega(t)\subset \mathbb{R}^d$ with boundary $\partial \Omega = \Gamma_0 \cup \Gamma_F$
split into a fixed (Dirichlet) and free surface part $\Gamma_0$ and $\Gamma_F$, respectively.
For the sake of economy we neglect inhomogeneous stress boundary conditions, though it would be straightforward to include them.
The domain at time $t$ is defined by advecting a reference configuration $\Omega_0$ by a mesh velocity $\mathbf{u}_{\mathrm{mesh}}(t)$
\begin{equation}
 \Omega(t) = \Omega_0 + \int_{t_0}^t \mathbf{u}_{\mathrm{mesh}}(t) \text{ d}t.
\end{equation}
The mesh velocity in normal direction (with unit normal $\mathbf{n}$) is given by the velocity $\mathbf{u}(t)$ on the free surface
\begin{equation}
 \mathbf{u}_\mathrm{mesh}(t) \cdot \mathbf{n} = \mathbf{u}(t) \cdot \mathbf{n} \quad \text{on} \quad \Gamma_F
\end{equation}
and set to zero on the rest of the boundary. We defer the discussion of the velocity $\mathbf{u}_\mathrm{mesh}$ in the interior of the domain
to Section~\ref{sec:remeshing}.
Finally, we denote the displacement at time $t$ by $\zeta = \int_{t_0}^t \mathbf{u}_\mathrm{mesh}(t)\cdot \mathbf{n} \text{ d}t$, leading to the evolution
equation
\begin{equation}
\frac{\text{d} \zeta}{\text{d}t} = \mathbf{u \cdot \mathbf{n}} \quad \textrm{on  }  \Gamma_F.
\label{eq:surface_evolution}
\end{equation}

@@

Equation~\eqref{eq:generalized_eigenvalue} is a generalized eigenvalue problem for the normal modes of the system.
It is rather more difficult to solve than a standard eigenvalue problem because the matrix on the right-hand-side 
is not invertible. It may, however, be transformed into a standard eigenvalue problem.
By defining
\begin{equation}
\begin{aligned}
C &= 
\begin{bmatrix}
A & B^T \\
B & 0 \\
\end{bmatrix} \quad
D &= 
\begin{bmatrix}
M & 0 \\
0 & 0
\end{bmatrix} \quad
\mathbf{y}_i &= 
\begin{bmatrix}
\mathbf{u}^n_i \\
p^n_i
\end{bmatrix} 
\end{aligned}
\end{equation}
and then multiplying both sides by $\tau_i^{-1}C^{-1}$ we get
\begin{equation}
(C^{-1}D)\mathbf{y}_i = \tau_i^{-1} \mathbf{y}_i
\label{eq:standard_eigenvalue}
\end{equation}

@@

Armed with the normal modes and relaxation times of the Stokes system, we can write down the
formal solution to Equation~\ref{eq:surface_evolution}. Let the initial surface topography be 
represented by a linear combination of its normal modes
\begin{equation}
\zeta(\mathbf{x}, t=0) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x})
\end{equation}
then the time evolution of the free surface is given by
\begin{equation}
\zeta( \mathbf{x}, t) = \displaystyle \sum_i a_i \zeta_i(\mathbf{x}) e^{-t/\tau_i}.
\end{equation}
Of course, most finite element (or finite difference, or finite volume) geodynamic simulations do not resolve 
the solution and surface topography into its normal modes and integrate those separately. 
Indeed, analytical solutions for the normal modes are only known for simple geometries and rheologies.
Instead, they obtain a velocity solution and simply advect the free surface with the local velocity.
The normal mode solution is instructive, however. 
Each mode has the form of a decay equation with characteristic decay time $\tau_i$.
The decay equation is the archetypical example of a stiff ordinary differential equation.
If we were to numerically integrate this in time with a forward Euler method, we would find the 
time-step criterion for stability \citep[e.g.][]{leveque2007finite} to be
\begin{equation}
\Delta t  \le 2 \tau_{\mathrm{min}},
\label{eq:cfl_euler}
\end{equation}
where $\tau_{\mathrm{min}}$ is this minumum decay time.
The maximum allowable timestep is limited by the minimum relaxation timescale.
If a larger timestep than this is taken then those modes will go unstable.
The modes with the smallest relaxation times are usually those with the largest lengthscales \citep{schubert2001mantle}, 
and so it will be those which go unstable first, a phenomenon which has been called 
a ``sloshing'', or ``drunken sailor'' instability \citep{kaus2010stabilization}.

@@

Again we may take one of the normal modes as an example and inspect the difference between the nonstandard finite-difference
scheme and the analytical solution after one timestep:

@@

\begin{itemize}
\item Direct solution of Equation~\eqref{eq:generalized_eigenvalue}. This can be expensive, although
only the minimun relaxation time is required. For certain geometries a simple power iteration on 
the standard eigenvalue problem~\eqref{eq:standard_eigenvalue} could be enough.
\item Analytical formulae. Several geometry and viscosity model combinations have analytical solutions
for relaxation spectra. Even if the user's model is not precisely the same (e.g., has some lateral viscosity
variations), an analytical approximation may be sufficient.
\item Scaling. In general, we expect the relaxation times to scale with $\tau \sim \frac{\eta}{\rho g L}$,
where the density, gravity, viscosity, and lengthscale $L$ are all representative values.  A small amount
of experimentation near this value of $\tau$ can find an appropriate relaxation time.
\item Observation of instabilities.  The mere act of observing a sloshing instability in an unstabilized
problem can furnish an estimate of its relaxation time.
\end{itemize}

@@

\subsection{Surface advection}
With a deformable domain there is the danger that small errors in free surface motion can
result in poor overall mass conservation in time. In some scenarios, the total volume of the mesh can 
fluctuate significantly over hundreds or thousands of timesteps.
Consistency with the Stokes solution requires 
\begin{equation}
\mathbf{u}_{\mathrm{mesh}} \cdot \mathbf{n} = \mathbf{u \cdot n}.
\end{equation}

@@

This solution is only valid for infinitesimal topography. However, for small
initial topography $\zeta_0$ it seems to be sufficiently accurate to test convergence orders up 
to quadratic \citep{kramer2012implicit, furuichi2015implicit}.
We estimate the error by time-integrating the L2 difference between the numerical and analytical
solutions at the center point of the free surface:
\begin{equation}
\mathrm{Error} = \frac{1}{4\tau}\int_0^{4 \tau} \lVert \zeta_\mathrm{numeric}(L/2, t) - \zeta_\mathrm{analytic}(L/2, t) \rVert_2 \text{ d}t.
\end{equation}

@@

\begin{figure*}
\includegraphics[width=0.9\textwidth]{figures/amr.pdf}
\caption{Convergence with degrees of freedom (DoFs) for the \citet{crameri2012comparison} Case 2 benchmark, for both uniform and adaptive mesh refinement. The timestep $\Delta t$ is 500 yr. We compare the maximum topography at 3 Myr with its value at high resolution ($\sim$398 m). For the adaptive cases we perform mesh refinement according to the sum of density and composition gradients. Quasi-implicit and nonstandard finite difference timestepping converge similarly, but the adaptive mesh refinement simulations require significantly smaller system sizes.}
\label{fig:amr}
\end{figure*}

@@

Finally, we have described the implementation of free surface boundary conditions in 
the open source mantle convection software \texttt{ASPECT}. Both the quasi-implicit 
scheme and the nonstandard finite difference scheme are available. The implementation is 
sufficiently general to accommodate many different geometries, rheologies, and 
gravity models. Furthermore, it runs in parallel and with adaptive mesh refinement.

\section*{Acknowledgements}
This work was supported through National Science Foundation grant EAR-1246670
and by the Computational Infrastructure in Geodynamics (CIG), through National Science Foundation grant EAR-0949446. 
TH is partially supported by NSF grant DMS-1522191.
Thanks to Wolfgang Bangerth, Brent Delbridge, and Cedric Thieulot for useful discussions.
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix
